--- 
# title: "Earth System Modelling"
# author: "Dr. Christian Seiler"
# date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
description: "ESM"
---

```{r, eval = TRUE, echo = FALSE}
.libPaths("/global/home/hpc5112/Rlibs_ESM")
styler::style_dir()
knitr::opts_chunk$set(fig.path = "figure-html/")
```

This website is under development

# Course Overview

| Week | Earth system focus                             | Modelling focus                                                                                                                           |
| ---- | ---------------------------------------------- | ----------------------------------------------------------------------------------------------------------------------------------------- |
| 1    | Building a conceptual Earth System Model   | From diagrams to equations: reservoirs, fluxes, state variables, prognostic vs. diagnostic variables, conservation, numerical integration |
| 2    | The atmosphere: Earth's energy balance     | Time scales, forcing and response, equilibrium vs. transient response, spin-up, climate sensitivity                                       |
| 3    | Ocean circulation                          | Dynamical systems: equilibria, stability, Jacobian, eigenvalues, phase portraits                                                          |
| 4    | Ocean circulation under freshwater forcing | Nonlinear systems, feedbacks, multiple equilibria, hysteresis, tipping points                                                             |
| 5    | Ocean carbon chemistry                     | Coupling components, flux exchange, conservation in coupled models, tracers                                                               |
| 6    | Marine biology                             | Nonlinear interactions, emergent behaviour, oscillations, feedbacks, increasing model complexity                                          |
| 7    | Land carbon cycle                          | Parameters, uncertainty, sensitivity analysis, parameter space exploration                                                                |
| 8    | Cryosphere                                 | Threshold behaviour, positive feedbacks, slow components, timescale separation                                                            |
| 9    | Coupling the Earth System                  | Model architecture, modularity, component coupling, numerical stability, coupled-model spin-up                                            |
| 10   | Future climate experiments                 | Experimental design, scenarios, ensembles, uncertainty propagation, global sensitivity analysis                                           |
| 11   | Learning from observations                 | Model evaluation, calibration, parameter optimization, inverse modelling                                                                  |
| 12   | Student modelling investigations           | Models as scientific experiments: hypothesis testing, interpretation, communication                                                       |

##  Assessments

|Assignment | Type | Weight | Due Date|
|:----------|:-----|:-------|:--------|
| Tutorial Report | Individual | 50% | Tue, Dec 3, 2024|
| Project Presentation | Group | 10% | Fri, Mar 21, 2025|
| Project Report | Group | 40% | Mon, Apr 7, 2025|

## Contact

* Instructor: Dr. Christian Seiler
* Email: christian.seiler@queensu.ca
* Office: Biosciences Complex, Room 3240, 116 Barrie Street
* Office hours: Tuesdays 2-4 PM

<!--chapter:end:index.Rmd-->

# CLIMBER-X

## Components
### Climate components

* 2.5-D semi-empirical statistical–dynamical atmosphere model (SESAM)
* 3-D frictional–geostrophic ocean model (GOLDSTEIN)
* Dynamic–thermodynamic sea ice model (SISIM)
* Land surface model (PALADYN)

### Optional components
* Ocean biogeochemistry and marinesediment model (HAMOCC)
* Ice sheet model for land (SCIPOLIS and YELMO)
* Viscoelastic mantle model (VILMA)

## Model structure
* The climate components are coupled through fluxes of energy and water, which are conserved
* Momentum is not conserved, but surface wind stress is computed in the atmosphere and passed to the ocean component
* Atmosphere, ocean, sea ice and land all have a horizontal resolution of 5 x 5 degrees
* Grid cells have land–sea fractions
* Base time step: 1 day (same as coupling frequency)
* The model does not represent interannual internal climate variability such as ENSO
* The land–ocean distribution, topography and land cover can evolve with time. A
* FORTRAN 90+
* Parallelized using OpenMP
* One namelist file for each model component

## Atmosphere model: SESAM

* The spatial resolution of SESAM is 2.5-D. This means that although prognostic variables (temperature, specific humidity, dust and eddy kinetic energy) as well as many diagnostic characteristics (cloudiness, condensation, shortwave radiation, etc.) are 2-D (latitude and longitude), the calculation of horizontal energy and water transport and the vertical fluxes of longwave radiation make use of information about the 3-D distribution of temperature, humidity, and wind velocity. All the vertical variations are purely diagnostic in the model.
* At the surface, each model grid cell is divided into different macro surface types, namely ocean, sea ice, land, lakes, and ice sheets. 
* The macro surface types share the same atmospheric column but differ in surface elevation, albedo, and surface temperature and humidity. 
* Sensible heat flux, evaporation, and shortwave and longwave radiation fluxes are computed separately for each macro surface type.
* The 3-D atmospheric structure is derived using assumptions about the universal vertical structure of temperature and relative humidity in the atmosphere. 
* The temperature profile through the troposphere is a quadratic function of height, except in a 1500 m layer close to the surface where the lapse rate is a function of the near-surface atmospheric stability
* Inversion strength is determined by the surface radiative balance and the difference between near-surface air temperature and skin temperature. 
* Temperature is vertically uniform in the stratosphere. 
* The tropopause height is derived assuming that the stratosphere is in radiative equilibrium. 
* Relative humidity is taken to be constant within the planetary boundary layer, to decay exponentially in the low troposphere and to remain constant throughout the rest of the troposphere. 
* Stratospheric relative humidity is prescribed at a uniform constant value.
* The horizontal wind velocity is divided into geostrophic and ageostrophic components. 
* The geostrophic components of velocity at sea level are computed from sea-level pressure (SLP) and at any height through the troposphere by using the thermal wind approximation
* The components of the surface wind are computed using the Taylor model

* The ageostrophic wind components in the planetary boundary layer are computed from the SLP gradient and the crossisobar angle. 
* SLP is the sum of zonally averaged and azonal components. 
* The zonal component of SLP is derived based on the assumption that the strength and position of the major cells of the atmospheric meridional overturning circulation are controlled by average meridional temperature gradients and zonally averaged surface drag

* The zonally averaged SLP is directly related to the zonal mean meridional (ageostrophic) wind in
the planetary boundary layer and therefore the mean meridional circulation. 

* The vertical structure of the zonally averaged meridional circulation is parameterized

* The azonal component of mean SLP resulting from stationary planetary waves is separated into a thermal component and an orographic component. 

* The thermal component is assumed to be proportional to the azonal surface temperature anomaly reduced
to sea level using the interrelation between long-term largescale azonal temperature and pressure in quasi-stationary planetary-scale waves

* SESAM includes a prognostic equation for vertically integrated energy and another for total-column water content, from which surface air temperature and humidity are derived.

* Sources and sinks of energy in the column include net longwave and shortwave radiation, surface sensible heat flux and latent heat release from condensation. 

* Sources and sinks of water in the column are given by surface evaporation and precipitation. 

* The 3-D wind field is used to advect the 3-D potential temperature field and 3-D specific humidity fields.













<!--chapter:end:00-climber-x_overview.Rmd-->

# Running CLIMBER-X

## Model configuration

* The main model configuration is defined in 

```{bash, eval=FALSE, echo=TRUE}
/CLIMBER-X/climber-x/nml/control.nml
```

* Here you find:

```{bash, eval=FALSE, echo=TRUE}
 flag_co2         = F        ! use interactive co2?
 flag_ch4         = F        ! use interactive ch4?
 flag_n2o         = F        ! use interactive n2o?
 flag_atm         = T        ! use atmosphere model?
 flag_ocn         = T        ! use ocean model?
 flag_bgc         = F        ! use ocean biogeochemistry model?
 flag_sic         = T        ! use sea ice model?
 flag_lnd         = T        ! use land model?
 flag_dust        = T        ! use dust model?
 flag_lakes       = F        ! use lake model?
 flag_geo         = F        ! use interactive geography/topography?
 flag_smb         = F        ! use surface mass balance model?
 flag_bmb         = F        ! use basal mass balance model?
 flag_ice         = F        ! use ice sheet model?
 ice_model_name   = 'yelmo'  ! select which ice sheet model to use, 'sico' or 'yelmo'
 ice_domain_name  = 'NH-32KM' !'ANT-32KM'      ! name of ice sheet domanin(s) to be used
```


## Simulations

* historical evaluation
* abrupt 4 x CO2
* doubling CO~2~
* Vegetation feedbacks
* Last glacial maximum
* freshwater hosing experiments
* Zero Emissions Commitment Model Intercomparison Project (ZECMIP)
* SSP scenarios


* Go to the directory where the CLIMBER-X code is located, e.g.

```{bash, eval=FALSE, echo=TRUE}
cd /CLIMBER-X/climber-x
```

* If you are running the model for the first time, create the runme configuration file from the template

```{bash, eval=FALSE, echo=TRUE}
cp .runme/runme_config .runme_config
```

* Edit the ```.runme_config``` as below. Replace ```your.email@queensu.ca``` with your email address:

```{bash, eval=FALSE, echo=TRUE}
{
    "hpc"           : "Frontenac",
    "account"       : "def-hpcg2003", 
    "jobname"       : "ClimberX",
    "omp"           : 32,
    "mem"           : -1,

    "queues_file"   : ".runme/queues.json",
    "info_file"     : ".runme/climberx_info.json",

    "email"         : "your.email@queensu.ca",
    "mail_type"     : ["FAIL", "REQUEUE"],
    "qos"           : "normal"

}
```

## Setting up an experiment

* Here is a minimum example for submitting a run:

```{bash, eval=FALSE, echo=TRUE}
./runme -rs -o RUNDIR
```

* where ```RUNDIR``` is the output directory, ```-r``` means that the job should actually be run and ```-s``` specifies that the job should be run on the cluster

* You can change parameter values using the ```-p``` option, e.g.:

```{bash, eval=FALSE, echo=TRUE}
./runme -rs -o RUNDIR -p ctl.n_accel=10
```

* ```ctl``` refers to the namelist group control as defined in ```.runme/climberx_info.json```. This json file looks as follows:

```{bash, eval=FALSE, echo=TRUE}
{

    "exe_default" : "climber",

    "par_path_as_argument" : false,

    "exe_aliases" : 
        {   "climber" : "climber.x",
            "mapping" : "mapping.x"
        },
    
    "grp_aliases" :
        {   "ctl"  : "control",
            "atm"  : "atm_par",
            "bgc"  : "bgc_par",
            "geo"  : "geo_par",
            "bmb"  : "bmb_par",
            "lnd"  : "lnd_par",
            "ocn"  : "ocn_par",
            "sic"  : "sic_par",
            "smb"  : "smb_par",
            "hose" : "hyster_hosing",
            "icey" : "ice_yelmo_par",
            "ices" : "ice_sico_par"
            
        },

    "par_paths" :
        {   "climber"  : [ "nml/control.nml",
                           "nml/atm_par.nml",
                           "nml/lnd_par.nml",
                           "nml/sic_par.nml",
                           "nml/bgc_par.nml",
                           "nml/geo_par.nml",
                           "nml/bmb_par.nml",
                           "nml/ocn_par.nml",
                           "nml/smb_par.nml",
                           "nml/ice_yelmo_par.nml",
                           "nml/ice_sico_par.nml",
                           "nml/hyster_ctrl.nml" ]
        },

    "files" : 
        ["nml/ice_grids.nml"], 

    "dir-special" :
    {
        "restart/vilma" : "vilma_restart"
    },

    "links" : 
        ["input", "maps", "restart"]

}
```

* So, ```ctl``` refers to ```control```. The control parameters are defined in the control namelist file:

```{bash, eval=FALSE, echo=TRUE}
/CLIMBER-X/climber-x/nml/control.nml
```

* For instance, the ``` n_accel``` is defined in ```control.nml``` as:

```{bash, eval=FALSE, echo=TRUE}
 n_accel = 1 ! acceleration factor of climate to ice sheet model
```

## Runme options

* Type ```runme -h``` to see all options:

```
[cseiler@rorqual1 climber-x]$ ./runme -h
usage: runme [-h] [-e EXE] [-r] [-s] [-q QUEUE] [-w WALL] [--part PARTITION] [--qos QOS] [-m MEM] [--omp OMP] [--email EMAIL] [--account ACCOUNT] [-x] [-v] [-p KEY=VALUE [KEY=VALUE ...]]
             -o RUNDIR

options:
  -h, --help            show this help message and exit
  -e EXE, --exe EXE     Define the executable file to use here. Shortcuts: climber=climber.x; mapping=mapping.x
  -r, --run             Run the executable after preparing the job?
  -s, --submit          Run the executable after preparing the job by submitting to the queue?
  -q QUEUE, --queue QUEUE
                        Alias of the queue the job should be submitted to.
  -w WALL, --wall WALL  HPC wall time "hh:mm:ss" specification to allow for submitted jobs (overrides settings given by queue alias)
  --part PARTITION      HPC partition specification (overrides settings given by queue alias)
  --qos QOS             HPC quality of service specification (overrides settings given by queue alias)
  -m MEM, --mem MEM     HPC max memory per node specification (overrides settings given by queue alias)
  --omp OMP             Specify number of threads for running with OpenMP (default = 1 implies no parallel computation)
  --email EMAIL         Email address to send job notifications from cluster (overrides config settings).
  --account ACCOUNT     HPC account associated with job, often set for resource permissions (overrides config settings).
  -x                    This argument is used when the script called by jobrun
  -v                    Verbose script output?
  -p KEY=VALUE [KEY=VALUE ...]
                        Set a number of key-value pairs (do not put spaces before or after the = sign). If a value contains spaces, you should define it with double quotes: foo="this is a
                        sentence". Note that values are always treated as strings.

required named arguments:
  -o RUNDIR             Path where simulation will run and store output.
```


## Submitting an ensemble

To submit an ensemble of runs with different parameter vales:

```{bash, eval=FALSE, echo=TRUE}
jobrun ./runme -rs -o OUTDIR -p ctl.n_accel=1,5,10
```

* where the individual runs will all be stored in ```OUTDIR```

<!--chapter:end:00-climber-x_running.Rmd-->

# Conceptual ESM

Main features:

* Two-latitude surface EBM (low/high)
* Deep ocean temperature
* Stommel circulation
* Salinity feedback
* Two-latitude upper-ocean carbon boxes with temperature-dependent air–ocean exchange
* Land carbon with temperature effects on photosynthesis and respiration, plus CO2 fertilization effect
* Atmospheric carbon
* Stefan–Boltzmann radiation, ice-albedo feedback, and lumped fast climate feedbacks (lapse rate, water vapor, clouds)

State Variables

| Variable       | Description                              |
|----------------|------------------------------------------|
| $T_L(t)$       | Low-latitude surface temperature (K)     |
| $T_H(t)$       | High-latitude surface temperature (K)    |
| $T_D(t)$       | Deep ocean temperature (K)               |
| $\Delta S(t)$  | Salinity contrast (psu)                  |
| $Q(t)$         | Thermohaline circulation strength (conceptual units) |
| $C_{atm}(t)$   | Atmospheric carbon (PgC)                 |
| $C_L(t)$       | Low-latitude upper ocean carbon (PgC)    |
| $C_H(t)$       | High-latitude upper ocean carbon (PgC)   |
| $C_{land}(t)$  | Land carbon (PgC)                        |
| $E(t)$         | Anthropogenic emissions (PgC/yr)         |

Surface Energy Balance - Low Latitude:

$$
H_L \frac{dT_L}{dt} =
S_L (1 - \alpha_L) 
- \varepsilon \sigma T_L^4
- D (T_L - T_H)
- \gamma Q (T_L - T_H)
+ \lambda F_{CO2}(C_{atm})
$$

Surface Energy Balance — High Latitude

$$
\alpha_H(T_H) = \alpha_{ice} + \frac{\alpha_{water}-\alpha_{ice}}{2} \left[ 1 + \tanh\left(\frac{T_H - T_c}{\Delta_T}\right) \right]
$$

$$
H_H \frac{dT_H}{dt} =
S_H (1 - \alpha_H(T_H))
- \varepsilon \sigma T_H^4
+ D (T_L - T_H)
+ \gamma Q (T_L - T_H)
- \kappa (T_H - T_D)
+ \lambda F_{CO2}(C_{atm})
$$

Deep Ocean Temperature

$$
H_D \frac{dT_D}{dt} = \kappa (T_H - T_D)
$$

Thermohaline Circulation (Stommel)

$$
\frac{dQ}{dt} = \lambda_Q \left[ \alpha_T (T_L - T_H) - \alpha_S \Delta S - \mu Q \right]
$$

Salinity Contrast

$$
\frac{d\Delta S}{dt} = F_{fw}(T_H) - Q \Delta S
$$

$$
F_{fw}(T_H) = F_0 + \eta (T_H - T_{ref})
$$

Ocean Carbon — Low Latitude Box

$$
\frac{dC_L}{dt} = k_{ao,L}(T_L) \left( C_{atm} - C_L \right) - Q (C_L - C_H)
$$

Ocean Carbon — High Latitude Box

$$
\frac{dC_H}{dt} = k_{ao,H}(T_H) \left( C_{atm} - C_H \right) + Q (C_L - C_H) - k_{oc} (C_H - C_{deep})
$$

Land Carbon

$$
\frac{dC_{land}}{dt} = P(T_L, C_{atm}) - R(T_L) C_{land}
$$

$$
P(T_L, C_{atm}) = P_0 f(T_L) \left( 1 + \phi \ln \frac{C_{atm}}{C_{ref}} \right) \\
R(T_L) = R_0 \exp[ \gamma_R (T_L - T_{ref}) ] \\
f(T_L) = \exp\left[ - \frac{(T_L - T_{opt})^2}{\Delta_T^2} \right]
$$
Atmospheric Carbon

$$
\frac{dC_{atm}}{dt} =
- k_{ao,L}(T_L) (C_{atm} - C_L)
- k_{ao,H}(T_H) (C_{atm} - C_H)
- P(T_L, C_{atm}) + R(T_L) C_{land} + E(t)
$$

Radiative Forcing

$$
F_{CO2}(C_{atm}) = \xi \ln \frac{C_{atm}}{C_{ref}}
$$

Temperature-dependent air-ocean exchange

$$
k_{ao,L}(T_L) = k_{ao,0} [ 1 - \beta (T_L - T_{ref}) ] \\
k_{ao,H}(T_H) = k_{ao,0} [ 1 - \beta (T_H - T_{ref}) ]
$$

Parameter Values

| Parameter      | Meaning                              | Value                                     |
|----------------|--------------------------------------|-------------------------------------------|
| $H_L$          | Surface heat capacity, low lat       | $1.0\times10^8$ J m$^{-2}$ K$^{-1}$     |
| $H_H$          | Surface heat capacity, high lat      | $1.0\times10^8$                          |
| $H_D$          | Deep ocean heat capacity             | $5.0\times10^9$                          |
| $S_L$          | Insolation low lat                   | 340 W m$^{-2}$                            |
| $S_H$          | Insolation high lat                  | 280 W m$^{-2}$                            |
| $\alpha_L$     | Low-lat albedo                        | 0.3                                       |
| $\alpha_{ice}$ | High-lat ice albedo                   | 0.6                                       |
| $\alpha_{water}$ | High-lat water albedo               | 0.3                                       |
| $T_c$          | Ice transition temp                   | 263 K                                     |
| $\Delta_T$     | Albedo transition width               | 5 K                                       |
| $\varepsilon$  | Emissivity                            | 1.0                                       |
| $\sigma$       | Stefan-Boltzmann                      | $5.67\times10^{-8}$ W m$^{-2}$ K$^{-4}$  |
| $D$            | Atmospheric heat transport           | 1.0 W m$^{-2}$ K$^{-1}$                   |
| $\gamma$       | Ocean heat transport                  | 0.5 W m$^{-2}$ K$^{-1}$                   |
| $\kappa$       | Deep ocean coupling                   | 0.1 W m$^{-2}$ K$^{-1}$                   |
| $\lambda$      | Climate sensitivity factor            | 0.8 (W m$^{-2}$)$^{-1}$                   |
| $\xi$          | CO$_2$ forcing coefficient            | 5.35 W m$^{-2}$                           |
| $\alpha_T$     | Thermal contribution to $Q$           | 0.1                                       |
| $\alpha_S$     | Salinity contribution to $Q$          | 0.08                                      |
| $\mu$          | Circulation damping                   | 0.2                                       |
| $\lambda_Q$    | Circulation timescale                 | 0.5 yr$^{-1}$                             |
| $F_0$          | Baseline freshwater                   | 0 psu/yr                                  |
| $\eta$         | Freshwater temp sensitivity           | 0.01 psu/K                                |
| $T_{ref}$      | Reference temperature                 | 288 K                                     |
| $k_{ao,0}$     | Base air-ocean carbon exchange        | 0.3 yr$^{-1}$                             |
| $\beta$        | Temp dependence exchange              | 0.005 K$^{-1}$                            |
| $k_{oc}$       | Ocean to deep ocean exchange          | 0.05 yr$^{-1}$                            |
| $P_0$          | Base photosynthesis                   | 90 PgC/yr                                 |
| $T_{opt}$      | Optimum photosynthesis temp           | 293 K                                     |
| $\phi$         | CO$_2$ fertilization                  | 0.2                                       |
| $R_0$          | Base respiration                      | 0.02 yr$^{-1}$                            |
| $\gamma_R$     | Respiration temp sensitivity          | 0.08 K$^{-1}$                             |
| $C_{ref}$      | Reference carbon                      | 280 PgC                                   |

<!--chapter:end:00-conceptual-ESM.Rmd-->

# Energy Balance

Student Name: ___

```{r, include = FALSE}
# template code chunk

student_id <- "12345678" # Replace with your Student ID

.libPaths("/global/home/hpc5112/Rlibs_ESM")
styler::style_dir()

library(jsonlite)

answers <- list()
answerKeys <- list()

record_answer <- function(value) {
  id <- knitr::opts_current$get("label")
  answers[[id]] <<- value
}

record_answerKey <- function(value) {
  id <- knitr::opts_current$get("label")
  answerKeys[[id]] <<- value
}
```

## Budyko–Sellers type EBM

## Carbon cycle

$$
C \frac{dT}{dt} = (1 - \alpha(T)) Q - \epsilon \sigma T^4
$$

$$
\alpha(T) = 0.5 - 0.2 \tanh \left( \frac{T-265}{10} \right) 
$$

* $Q$ is the mean incoming solar radiation: $Q = \frac{1}{4} S_0$, where $S_0$ = 1368 W m$^{-2}$ is the solar constant, such that $Q$ = 342 W m$^{-2}$
* $C$ = planetary heat capacity ($C$ = 5 $\times$ 10$^8$ J m$^{-2}$ K$^{-1}$) [Schwartz (2007)](https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2007JD008746)


```{r, fig.width = 5, fig.height = 5}
T <- seq(220, 310, 1)

# Outgoing radiation following the Stefan-Boltzmann law
emissivity <- 0.6
sigma <- 5.67 * 10^(-8) # Stefan-Boltzmann constant in W m-2 K-4
E_out <- emissivity * sigma * T^4

# Incoming radiation
Q <- 1368 / 4 # mean incoming solar radiation in W m-2
albedo <- 0.5 - 0.2 * tanh((T - 265) / 10)
E_in <- (1 - albedo) * Q

plot(T, E_out,
  type = "l", col = "red",
  xlab = "Temperature (K)",
  ylab = "Radiative Flux (W/m2)"
)
lines(T, E_in, col = "blue")
grid()
legend("topleft", c("Incoming radiation", "Outgoing radiation"), pch = 16, col = c("blue", "red"), bty = "n")
```



```{r, fig.width = 5, fig.height = 5}
Q <- 1368 / 4 # mean incoming solar radiation in W m-2
# C <- 5 * 10^8 # planetary heat capacity in J m$^{-2}$ K$^{-1}$) [Schwartz (2007)](https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2007JD008746)
emissivity <- 0.6
sigma <- 5.67 * 10^(-8) # Stefan-Boltzmann constant in W m-2 K-4

T <- seq(220, 310, 1)
albedo <- 0.5 - 0.2 * tanh((T - 265) / 10)
CdTdt <- (1 - albedo) * Q - emissivity * sigma * T^4

plot(T, CdTdt, type = "l", xlab = "Temperature (K)", ylab = "Energy Flux (W/m2)")
grid()
abline(h = 0)
```


```{r, fig.width = 5, fig.height = 5, eval = TRUE, echo = FALSE, warning = FALSE}
# template code chunk

# ----------------------------
# Parameters
# ----------------------------

Q0 <- 1368 / 4
emissivity <- 0.6
sigma <- 5.67e-8

# ----------------------------
# Equilibrium function
# ----------------------------

Ffun <- function(T, mu) {
  Q <- mu * Q0
  albedo <- 0.5 - 0.2 * tanh((T - 265) / 10)
  result <- (1 - albedo) * Q - emissivity * sigma * T^4
  return(result)
}

# ----------------------------
# Parameter sweep
# ----------------------------

mu_vals <- seq(0.7, 1.4, length.out = 300)

T_all <- c()
mu_all <- c()

T_grid <- seq(200, 330, by = 0.5)

for (mu in mu_vals) {
  Fvals <- Ffun(T_grid, mu)
  sign_change <- which(diff(sign(Fvals)) != 0)

  if (length(sign_change) > 0) {
    for (idx in sign_change) {
      root <- uniroot(Ffun,
        interval = c(T_grid[idx], T_grid[idx + 1]),
        mu = mu
      )$root

      T_all <- c(T_all, root)
      mu_all <- c(mu_all, mu)
    }
  }
}

# ----------------------------
# Plot bifurcation diagram
# ----------------------------

plot(mu_all, T_all,
  pch = 16,
  cex = 0.6,
  xlab = expression(Q / Q[0]),
  ylab = "Equilibrium Temperature (K)"
)

grid()
```


<!--chapter:end:00-energy-balance.Rmd-->


# Global Carbon Cycle

$$
c_1\frac{dT_1}{dt} = \lambda T_1 + \kappa(T_2 - T_1) + Q_2 \ln\frac{C_A}{C^*_A}
$$
$$
c_2\frac{dT_2}{dt} = -\kappa(T_2 - T_1)
$$

$$
\frac{dC_L}{dt} = \Pi C_A - \Pi C_A^*\frac{C_L}{C_L^*} \exp(\alpha \beta T_1)
$$

$$
\frac{dC_1}{dt} = \nu_1 \left( C_A - \epsilon(C_1) \right) - \nu_2 \left( C_1 - \frac{V_1}{V_2} C_2 \right)
$$
$$
\frac{dC_2}{dt} = \nu_2 \left( C_1 - \frac{V1}{V2} C_2 \right)
$$

$$
\Pi(C_A) = \Pi C_A^*
$$
* $c_1$ upper box heat capacity ($c_1$ = 7.7 W a m$^{-2}$ K$^{-1}$)
* $c_1$ lower box heat capacity ($c_2$ = 100 W a m$^{-2}$ K$^{-1}$)
* $T_1$ is the globally averaged temperature anomaly of the atmosphere and upper ocean in K
* $T_2$ is the globally averaged temperature anomaly of the deep ocean in K
* $\lambda$ is the climate feedback parameter
* $\kappa$ is a heat transfer coefficient ($\kappa$ = 0.57 W m$^{-2}$ K$^{-1}$)
* $Q_2$ Radiative forcing due to doubled CO2 (3.7 W m$^{-2}$) (Dessler: uses 5.35 W m$^{-2}$)
* $C_A$ is the atmospheric carbon in ppmv
* $C^*_A$ is the PI atmospheric carbon in ppmv ($C^*_A = 286$ ppmv)
* $V_2/V_1$ is the ocean box volume ratio ($V_2/V_1 = 50$)




<!--chapter:end:00-global-carbon-cycle.Rmd-->

::: {.include}
# Topic

```{r, include = FALSE}
# Add your Name and Student ID here:
Name <- "Christian Seiler"
ID <- "12345678"
print(paste("Student Name:", Name, ", Student ID:", ID))
```

```{r, include = FALSE}
# Do not alter this code chunk

.libPaths("/global/home/hpc5112/Rlibs_ESM")
styler::style_dir()

library(jsonlite)

answers <- list()
answerKeys <- list()

record_answer <- function(value) {
  id <- knitr::opts_current$get("label")
  answers[[id]] <<- value
}

record_answerKey <- function(value) {
  id <- knitr::opts_current$get("label")
  answerKeys[[id]] <<- value
}
```
:::

Some random text.

::: {.question answer="A"}

**Q: The two dots denote conditions representative for the low-latitude and high-latitude boxes. Which one is which?**

* (A): low-latitude = red, high-latitude = blue
* (B): low-latitude = blue, high-latitude = red

```{r, echo = FALSE}
answer <- "X"
record_answer(answer)
print(answer)
```
:::

More random text.

```{r, include=FALSE}
tutorialName <- "ocean-circulation"

output <- list(
  student_id = student_id,
  tutorial   = tutorialName,
  timestamp  = Sys.time(),
  answers    = answers
)

filename <- paste0("answers_", tutorialName, "_", student_id, ".json")
jsonlite::write_json(output, filename, pretty = TRUE)
rm(output)
```

```{r generate_solution_JSON, include=FALSE}
# Generate solution JSON file for this tutorial

library(jsonlite)

# Build the answer key object
output <- list(
  tutorial = tutorialName,
  generated_at = Sys.time(),
  answers = answerKeys
)

# Write JSON file
filename <- paste0("answerKey_", tutorialName, ".json")
jsonlite::write_json(output, filename, pretty = TRUE)
```


<!--chapter:end:00-topic.Rmd-->

# Glacial Oscillations

Saltzman and Maasch (1990) (SM90) developed a conceptual model to simulate glacial cycles during the late Cenozoic era (300 kyr BCE). The ODEs below present the nondimensional form of the model, where $x$, $y$, and $z$ are nondimensional variables representing global ice mass, atmospheric CO$_2$ concentration, and deep ocean temperature, respectively:

$$ \frac{dx}{dt} = -x - y - vz, $$
$$ \frac{dy}{dt} = - pz + ry - sy^2 - y^3, $$
$$ \frac{dz}{dt} = -qx -qz, $$
where $p$ = 1, $q$= 2.5, $r$ = 1.3, $s$ = 0.6, and $v$ = 0.2. 

Note that this version of the model ignores changes in incoming solar radiation in response to the Milankovitch cycle and is intended for studying the internal oscillation of the system. Let us begin by plotting the free solution of the dynamical system.

```{r, fig.width = 5, fig.height = 5}
# Load required package
library(deSolve)

# Define parameters (change these to explore different behavior)
p <- 1
q <- 2.5
r <- 1.3
s <- 0.6
v <- 0.2

# Define the system of ODEs
system <- function(t, state, parameters) {
  x <- state[1]
  y <- state[2]
  z <- state[3]

  dx <- -x - y - v * z
  dy <- -p * z + r * y - s * y^2 - y^3
  dz <- -q * (x + z)

  list(c(dx, dy, dz))
}

# Initial conditions
initial_conditions <- c(0.1, 0.8, -0.2)

# Time span

dt <- 0.01
times <- seq(0, 100, by = dt)

# Solve ODEs for each initial condition
solutions <- ode(
  y = initial_conditions,
  times = times,
  func = system,
  parms = NULL,
  method = "lsoda"
)

# Plot
df <- data.frame(solutions)
colnames(df) <- c("time", "x", "y", "z")

plot(
  x = df$time,
  y = df$x,
  type = "l",
  ylim = c(-2, 2),
  ylab = "(-)",
  xlab = "Time (10 ky)",
  xlim = c(0, 30)
)
lines(x = df$time, df$y, col = "red")
lines(x = df$time, df$z, col = "blue")
abline(h = 0, lty = 2)
legend(
  "top",
  c("x = Ice", "y = CO2", "z = T_Ocean"),
  col = c("black", "red", "blue"),
  pch = 16,
  bty = "n",
  horiz = TRUE
)
```


Impact of Milankovitch cycle

```{r}
library(palinsol)

year <- seq(from = -500000, to = 0, by = 1000)

latitude <- 65 * pi / 180 # (radians)

insolation <- function(times, astrosol = ber78) {
  sapply(times, function(tt) Insol(orbit = astrosol(tt), long = pi / 2, lat = latitude))
}

# Daily mean incoming solar radiation at TOA (W/m2)
isl <- insolation(year, ber78)

# Normalized anomalies

isl.norm <- (isl - mean(isl)) / sd(isl)

my.min <- min(isl.norm)
my.max <- max(abs(isl.norm))

Rvec <- (isl.norm / my.max)

plot(year, isl,
  typ = "l",
  main = "Daily mean solar radiation for June 20 (Solstice)",
  xlab = "Years",
  ylab = "Insolation at TOA (W/m2)"
)
legend("topleft", paste("Latitude =", latitude * 180 / pi, "N"), bty = "n")
abline(h = 480.05, col = "grey")

plot(year, Rvec,
  typ = "l",
  main = "Non-dimensionalized insolation R (-)",
  xlab = "Years",
  ylab = "(-)"
)
legend("topleft", paste("Latitude =", latitude * 180 / pi, "N"), bty = "n")
```

Le

```{r, fig.width = 5, fig.height = 5}
# Load required package
library(deSolve)

# Define parameters (change these to explore different behavior)
p <- 1
q <- 2.5
r <- 1.3
s <- 0.6
v <- 0.2
u <- 0.5

# Define the system of ODEs
system <- function(t, state, parameters) {
  x <- state[1]
  y <- state[2]
  z <- state[3]

  k <- floor(t / dt) + 1
  Rt <- Rvec[k]

  dx <- -x - y - v * z - u * Rt
  dy <- -p * z + r * y - s * y^2 - y^3
  dz <- -q * (x + z)

  list(c(dx, dy, dz))
}

# Initial conditions
# initial_conditions <- c(0.1, 0.8, -0.2)
# initial_conditions <- c(0.5, -0.8, -0.2)
initial_conditions <- c(1, -1, -1)

# Time span

dt <- 1 / 10
times <- seq(0, 50, by = dt)

# Solve ODEs for each initial condition
solutions <- ode(
  y = initial_conditions,
  times = times,
  func = system,
  parms = NULL,
  method = "lsoda"
)

# Plot
df <- data.frame(solutions)
colnames(df) <- c("time", "x", "y", "z")

plot(
  x = df$time,
  y = df$x,
  type = "l",
  ylim = c(-2, 2),
  ylab = "(-)",
  xlab = "Time (10 ky)",
  xlim = c(0, 50)
)
lines(x = df$time, df$y, col = "red")
lines(x = df$time, df$z, col = "blue")
abline(h = 0, lty = 2)
legend(
  "top",
  c("x = Ice", "y = CO2", "z = T_Ocean"),
  col = c("black", "red", "blue"),
  pch = 16,
  bty = "n",
  horiz = TRUE
)
```

<!--chapter:end:01-intro.Rmd-->

# Linear Algebra

This section introduces the basic mathematical concepts required for understanding tipping points.

## Matrix Multiplication

$$
\begin{bmatrix}
1 & 2 \\
3 & 4
\end{bmatrix}
\begin{bmatrix}
5 \\
6
\end{bmatrix}
=
\begin{bmatrix}
1 \cdot 5 + 2 \cdot 6 \\
3 \cdot 5 + 4 \cdot 6
\end{bmatrix}
=
\begin{bmatrix}
17 \\
39
\end{bmatrix}
$$

## Matrix Transposition

$$
\mathbf{A} =
\begin{bmatrix}
1 & 2 \\
3 & 4 \\
5 & 6
\end{bmatrix}
\quad \Rightarrow \quad
\mathbf{A}^\top =
\begin{bmatrix}
1 & 3 & 5 \\
2 & 4 & 6
\end{bmatrix}
$$

## Matrix Inversion

For a square matrix $\mathbf{A}$, the inverse $\mathbf{A}^{-1}$ satisfies:

$$
\mathbf{A} \mathbf{A}^{-1} = \mathbf{A}^{-1} \mathbf{A} = \mathbf{I}
$$
where $\mathbf{I}$ is the identity matrix. Note that only non-singular (invertible) matrices have an inverse. For a $2 \times 2$ matrix:

$$
\mathbf{A} =
\begin{bmatrix}
a & b \\
c & d
\end{bmatrix}
\quad \Rightarrow \quad
\mathbf{A}^{-1} = \frac{1}{ad - bc}
\begin{bmatrix}
d & -b \\
-c & a
\end{bmatrix}
\quad \text{if } ad - bc \neq 0
$$

## Transformation

If we multiply a matrix with a vector, we apply a linear transformation to the vector. This produces a new vector. Depending on the transformation, the new vector may have new coordinates, a different length, and possibly a different direction. For example consider the matrix $\mathbf{A}$ and the vector $\mathbf{v}$:

$$
\mathbf{A} = \begin{bmatrix}
a & b \\
c & d
\end{bmatrix}
\text{ and }
\mathbf{v} = \begin{bmatrix}
x \\
y
\end{bmatrix}
$$
Let's multiply the two:

$$
\mathbf{Av} =
\begin{bmatrix}
a & b \\
c & d
\end{bmatrix}
\begin{bmatrix}
x \\
y
\end{bmatrix}
=
\begin{bmatrix}
ax + by \\
cx + dy
\end{bmatrix}
$$

The matrix transforms the vector. It produces a new vector whose length, direction, or coordinates may change. So matrices do not just store numbers, they describe how vectors and shapes are transformed. Consider the matrix $\mathbf{A}$ and the two vectors $\mathbf{v_1}$ and $\mathbf{v_2}$

$$
\mathbf{A} =
\begin{bmatrix}
2 & 1 \\
1 & 3 \\
\end{bmatrix}
\mathbf{v_1} =
\begin{bmatrix}
1 \\
0 \\
\end{bmatrix}
\mathbf{v_2} =
\begin{bmatrix}
0 \\
1 \\
\end{bmatrix}
$$
The two vectors span a unit square in the plane. Applying the matrix transforms that square into a parallelogram:

$$
\mathbf{Av_1} = 
\begin{bmatrix}
2 \\
1 \\
\end{bmatrix}
\mathbf{Av_2} = 
\begin{bmatrix}
1 \\
3 \\
\end{bmatrix}
$$


## Eigenvectors and eigenvalues

Consider the following transformation:

$$
\mathbf{Av} = \mathbf{v'}
$$

In general, $\mathbf{v'}$ points in a different direction than $\mathbf{v}$. But some vectors are special, they don't change the direction under the transformation. These special vectors are called eigenvectors. For these eigenvectors $\mathbf{v}$, the matrix $\mathbf{A}$ only stretches or shrinks them:

$$
\mathbf{Av} = \lambda \mathbf{v}
$$

* $\lambda$ is called the eigenvalue corresponding to that eigenvector 
* $|\lambda|>1$: vector is stretched
* $|\lambda|<1$: vector is shrunk
* $\lambda<0$: vector flips direction

For example, let's stretch a vector by a factor of two using a matrix:

$$
\mathbf{Av} =
\begin{bmatrix}
1 & 2 \\
1 & 0
\end{bmatrix}
\begin{bmatrix}
2 \\
1
\end{bmatrix}
=
\begin{bmatrix}
2 + 2 \\
2 + 0
\end{bmatrix}
=
\begin{bmatrix}
4 \\
2
\end{bmatrix}
= 2 \begin{bmatrix}
2 \\
1
\end{bmatrix}
= \lambda \mathbf{v}
$$
Here, $\mathbf{v}$ is the eigenvector, and the corresponding eigenvalue $\lambda$ is the factor by which the vector is being stretched ($\lambda = 2$). The eigenvalues can help us to understand how a system behaves close to its equilibrium points, as will be explained a little further below. 

## Determinant

A matrix transforms the shape spanned by vectors. The determinant tells us how the shape’s area changes, for example, whether it expands, shrinks, or collapses into a line. Consider the matrix:

$$
\mathbf{A} =
\begin{bmatrix}
a & b \\
c & d \\
\end{bmatrix}
$$

The determinant of matrix $\mathbf{A}$ is defined as:

$$
\det(\mathbf{A}) = ad -bc
$$
For example: 

$$
\mathbf{A} =
\begin{bmatrix}
1 & 2 \\
1 & 0 \\
\end{bmatrix}
$$

$$
\det(\mathbf{A}) = 1*0 - 2*1 = -2
$$

* $|\det({\mathbf{A}})|>1$: area expands
* $|\det({\mathbf{A}})|<1$: area shrinks
* $|\det({\mathbf{A}})|=0$: area collapses

In this case, the determinant tells us that applying the matrix $\mathbf{A}$ will expand the area spanned by vectors ($|-2|>1$). The determinant allows us to to calculate the eigenvalues $\lambda$:

$$
\mathbf{A} \mathbf{v} = \lambda \mathbf{v}
$$

* Let's rewrite this equation: $\mathbf{A} \mathbf{v} - \lambda \mathbf{v} = 0$
* Rearrange: $(\mathbf{A} - \lambda \mathbf{I}) \mathbf{v} = 0$
* Where $\mathbf{I}$ is the identity matrix:

$$
\mathbf{I} = 
\begin{bmatrix}
1 & 0 \\
0 & 1 \\
\end{bmatrix}
$$
If $\mathbf{v} \ne 0$, then:

$$
\det(\mathbf{A} - \lambda \mathbf{I}) = 0
$$
The determinant can therefore be used to calculate the eigenvalues. For instance:

$$
\mathbf{A} - \lambda \mathbf{I} = \begin{bmatrix} 2-\lambda & 3 \\ 1 & 4-\lambda \end{bmatrix}
$$

$$
\det(\mathbf{A} - \lambda \mathbf{I}) = (2-\lambda)(4-\lambda) - 3 \times 1
$$

$$
= (2-\lambda)(4-\lambda) - 3
$$

$$
= 2 \times 4 - 2\lambda - 4\lambda + \lambda^2 - 3 = 8 - 6\lambda + \lambda^2 - 3
$$

$$
= \lambda^2 - 6\lambda + 5
$$

$$
\lambda^2 - 6\lambda + 5 = 0
$$

$$
\lambda = \frac{6 \pm \sqrt{36 - 20}}{2} = \frac{6 \pm \sqrt{16}}{2} = \frac{6 \pm 4}{2}
$$

$$
\Rightarrow \quad \lambda_1 = 5, \quad \lambda_2 = 1
$$

## Jacobian Matrix

In dynamical systems theory, the matrix that we will be using is the Jacobian matrix $\mathbf{J}$. The Jacobian matrix is a matrix of partial derivatives that describes how a multivariable function changes locally. For instance:

$$
\mathbf{f}(\mathbf{x}) =
\begin{bmatrix}
f_1(x, y, z) \\
f_2(x, y, z) \\
f_3(x, y, z)
\end{bmatrix}
\mathbf{J} =
\begin{bmatrix}
\frac{\partial f_1}{\partial x} & \frac{\partial f_1}{\partial y} & \frac{\partial f_1}{\partial z} \\
\frac{\partial f_2}{\partial x} & \frac{\partial f_2}{\partial y} & \frac{\partial f_2}{\partial z} \\
\frac{\partial f_3}{\partial x} & \frac{\partial f_3}{\partial y} & \frac{\partial f_3}{\partial z}
\end{bmatrix}
$$

So the eigenvalue problem becomes: 

$$
\det(\mathbf{J} - \lambda \mathbf{I}) = 0
$$
The eigenvalues $\lambda_i$ of the Jacobian $\mathbf{J}$ determine the local stability and type of the equilibrium point. If all eigenvalues have negative real parts, the equilibrium is locally asymptotically stable (perturbations decay). If any eigenvalue has a positive real part, the equilibrium is unstable (perturbations grow). If eigenvalues have zero real parts, the system may be marginally stable.

## Numerical example

The goal is to identify and characterize equilibrium points. The main steps are:

* (1) Set the ODEs to zero and solving for each variable

* (2) Apply the Jacobian matrix to the ODEs

* (3) Calculate the eigenvalues

* (4) Calculate the equilibrium eigenvalues

The equilibrium eigenvalues ($\lambda$) characterize the equilibrium point, e.g.:

* If all $\lambda > 0$: unstable node
* If all $\lambda < 0$: stable node
* If the sign varies among $\lambda$: saddle (unstable)
* If any $\lambda = 0$: tipping point

### Step (1): Set the ODEs to zero and solving for each variable

Consider the following set of ordinary differential equations:

$$f_1: \frac{dx}{dt} = \mu - x^2 $$
$$f_2: \frac{dy}{dt} = -y $$

At equilibrium, $\frac{dx}{dt} = 0$ and $\frac{dy}{dt} = 0$. Find the values of $x$ and $y$ for which $f_1$ and $f_2$ equal zero. These values are denoted as $x^*$ and $y^*$.

* $0 = \mu - x^2$
* $\Leftrightarrow  x^2 = \mu$
* Therefore: $x^* = \pm \sqrt{\mu}$
* And $y^* = 0$

We conclude that

* For $\mu>0$ the equilibria are found at: $(\sqrt{\mu}, 0)$ and $(-\sqrt{\mu}, 0)$
* For $\mu=0$ the equilibrium is found at: $(0, 0)$
* For $\mu<0$ there is no equilibrium point

### Step (2): Apply the Jacobian matrix to the ODEs

* Recall our ODES: $f_1: \frac{dx}{dt} = \mu - x^2 $ and $f_2: \frac{dy}{dt} = -y$

* We now apply the Jacobian matrix to these ODEs:

$$
\mathbf{J} =
\begin{bmatrix}
\frac{\partial f_1}{\partial x} & \frac{\partial f_1}{\partial y} \\
\frac{\partial f_2}{\partial x} & \frac{\partial f_2}{\partial y} \\
\end{bmatrix}
=
\begin{bmatrix}
-2x & 0 \\
0 & -1 \\
\end{bmatrix}
$$

### Step (3) Calculate the eigenvalues

* Recall that eigenvalues $\lambda$ satisfy $\det(\mathbf{J} - \lambda \mathbf{I}) = 0$ , where $\mathbf{I}$ is the identity matrix:

$$
\mathbf{I} = 
\begin{bmatrix}
1 & 0 \\
0 & 1 \\
\end{bmatrix}
$$

* Therefore:

$$
\mathbf{J} - \lambda \mathbf{I} = 
\begin{bmatrix}
-2x & 0 \\
0 & -1 \\
\end{bmatrix}
-
\begin{bmatrix}
\lambda & 0 \\
0 & \lambda \\
\end{bmatrix}
=
\begin{bmatrix}
-2x-\lambda & 0 \\
0 & -1-\lambda \\
\end{bmatrix}
$$

* Compute the determinant:

$$
\det(\mathbf{J} - \lambda \mathbf{I}) = (-2x-\lambda) (-1-\lambda)
$$

* Set the determinant equal to zero:

$$ (-2x-\lambda) (-1-\lambda) = 0 $$

* This is possible if $(-2x-\lambda)$ and/or $(-1-\lambda)$ equal zero:

$$(-2x-\lambda) = 0 \text{ and } -1-\lambda = 0$$

* Solving for $\lambda$ gives the two eigenvalues $\lambda_1$ and $\lambda_2$:

$$\lambda_1 = -2x \text{ and }\lambda_2 = -1$$

###  Step (4): Calculate the equilibrium eigenvalues

* At $(x^*, y^*) = (\sqrt{\mu}, 0)$:

$$\lambda_1 = -2\sqrt{\mu}<0 \text{ and }\lambda_2 = -1 < 0$$

* Here, both eigenvalues are negative. This is a stable node.

* At $(x^*, y^*) = (-\sqrt{\mu}, 0)$:

$$\lambda_1 = 2\sqrt{\mu} > 0 \text{ and }\lambda_2 = -1 < 0$$

* Here, we have one positive and one negative eigenvalue. This is a saddle (unstable) 

* At $(x^*, y^*) = (0, 0)$:

$$\lambda_1 = 0 \text{ and }\lambda_2 = -1 < 0$$

* Here, one eigenvalue is zero. This is a tipping point.

<!--chapter:end:02-linearAlgebra.Rmd-->

::: {.include}
# Ocean Circulation

```{r, include = FALSE}
# Add your Name and Student ID here:
Name <- "Christian Seiler"
ID <- "12345678"
print(paste("Student Name:", Name, ", Student ID:", ID))
```

```{r, include = FALSE}
# Do not alter this code chunk

.libPaths("/global/home/hpc5112/Rlibs_ESM")
styler::style_dir()

library(jsonlite)

answers <- list()
answerKeys <- list()

record_answer <- function(value) {
  id <- knitr::opts_current$get("label")
  answers[[id]] <<- value
}

record_answerKey <- function(value) {
  id <- knitr::opts_current$get("label")
  answerKeys[[id]] <<- value
}
```
:::

## The Stommel Model

Henry Stommel developed a conceptual model of the thermohaline circulation, which he published in a seminal [paper](https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.2153-3490.1961.tb00079.x) in 1961. The model consists of two idealized ocean boxes representing low-latitude warm water and high-latitude cold water. The circulation between the boxes is driven by density differences: denser water sinks in the high-latitude box, flows equatorward at depth, and returns poleward near the surface, conserving mass. The strength of the circulation increases with the density contrast between the boxes. Water density depends on both temperature and salinity: higher salinity increases density, whereas higher temperature decreases it.

In the low-latitude box, the temperature is warmer because the overlying atmosphere is warmer, and the salinity is higher because prescribed evaporation exceeds precipitation. In the high-latitude box, the temperature is colder because the overlying atmosphere is colder, and the salinity is lower because prescribed precipitation exceeds evaporation. The temperature in both boxes is strongly constrained by the atmosphere through restoring toward atmospheric temperatures, although exchange of water between the boxes can transport heat. The salinity, however, is directly affected by the circulation: stronger circulation transports salt between the boxes and tends to reduce the salinity contrast.

The flow is driven by density differences, which depend on both temperature and salinity with opposing effects: higher temperature decreases density, whereas higher salinity increases it. Thus, the low-latitude box is warm and salty, while the high-latitude box is cold and fresher.

Two types of circulation are possible. If density differences are dominated by temperature, the high-latitude box is denser than the low-latitude box, and water flows from high to low latitudes in the deep branch with a compensating surface return flow. This is the thermally dominated circulation. If density differences are dominated by salinity, the low-latitude box becomes denser, and the overturning circulation reverses, with deep flow toward high latitudes. This is the salinity-dominated circulation.

Because circulation transports salt between the boxes, stronger flow reduces the salinity contrast. This tends to stabilize thermally dominated circulation but weaken salinity-dominated circulation.

### Temperature - Salinity Diagram

To better understand this, let's assess how density is affected by temperature and salinity. The relationship is non-linear, and determined by the Thermodynamic Equation of Sea Water 2010 (TEOS-10). The Figure below gives the sea water density (blue contours in kg m^-3^) as a function of salinity and temperature. 


```{r, fig.width = 5, fig.height = 5, echo = TRUE}
library(gsw)
# Define ranges
T <- seq(0, 30, by = 0.1) # y-axis (temperature)
S <- seq(30, 40, by = 0.1) # x-axis (salinity)

# Compute density using outer
# outer(T, S) -> rows = T, cols = S
rho_matrix <- outer(T, S, function(T, S) gsw_rho(SA = S, CT = T, p = 0))

# Transpose so contour interprets correctly
rho_matrix <- t(rho_matrix)

# Plot
contour(
  x = S, y = T, z = rho_matrix,
  xlab = "Salinity (g/kg)",
  ylab = "Temperature (°C)",
  main = "T-S Diagram with TEOS-10 Density",
  levels = seq(500, 2000, by = 2),
  col = "blue", lwd = 1, labcex = 1.0
)

box.HL <- c(34, 2) # salinity and temperature of high-latitude box
box.LL <- c(36, 25) # salinity and temperature of high-latitude box
points(x = box.HL[1], y = box.HL[2], col = "blue", pch = 16) # high-latitude
points(x = box.LL[1], y = box.LL[2], col = "red", pch = 16) # low-latitude
grid()
```

::: {.include}

**Q: The two dots denote conditions representative for the low-latitude and high-latitude boxes. Which one is which?**

* (A): low-latitude = red, high-latitude = blue
* (B): low-latitude = blue, high-latitude = red

```{r, echo = FALSE}
answer <- "X"
record_answer(answer)
print(answer)
```
:::

::: {.answer}
```{r, include = FALSE}
answerKey <- "A"
record_answerKey(answerKey)
```
:::

::: {.include}

**Q: The blue dot has a higher density compared to the red dot. Is this due to the:**

* (A): salinity contrast
* (B): temperature contrast

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```
:::

```{r, include = FALSE}
answerKey <- "B"
record_answerKey(answerKey)
```

**Q: Let's reduce salinity of the blue dot by 6 g/kg. Which dot would show the higher density?**

* (A): red
* (B): blue

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "A"
record_answerKey(answerKey)
```

**Q: How would this affect the direction of flow in the lower branch?**

* (A): The flow would be directed from low to high latitudes
* (B): The flow would be directed from high to low latitudes

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "A"
record_answerKey(answerKey)
```

**Q: What is the relevance of this finding for the context of climate change?**

* (A): Climate change causes ice sheets to melt, reducing the salinity of North Atlantic seawater, decreasing water density, and strengthening the Atlantic Meridional Overturning Circulation.
* (B): Climate change causes ice sheets to melt, reducing the salinity of North Atlantic seawater, decreasing water density, and weakening the Atlantic Meridional Overturning Circulation.
* (C): Climate change causes ice sheets to melt, increasing freshwater input to the North Atlantic, increasing water density, and weakening the Atlantic Meridional Overturning Circulation.
* (D): Climate change causes ice sheets to melt, freshening the North Atlantic, but the resulting salinity change has no significant influence on the Atlantic Meridional Overturning Circulation because temperature is the dominant control on density.

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```



```{r, include = FALSE}
answerKey <- "B"
record_answerKey(answerKey)
```

In the Stommel model, we use a linearlized version of the equation of state:

$$
\rho = \rho_0 (1 - \alpha (T- T_0)  + \beta (S-S_0)
$$

where $\rho$ is the sea water density (kg m$^{-3}$), $T$ and $S$ are temperature and salinity anomalies, respectively and $\alpha$ is the thermal contraction ($\alpha = 1.5 \times 10^{-4}$ K) and $\beta$ is the saline expansion ($\beta = 8 \times10^{-4}$ psu^-1^). The unit psu stands for practical salinity unit, which is a dimensionless unit that expresses the ratio of conductivities. A value of 1 psu roughly correspond 1 g salt per 1 kg sea water. $T_0$ and $S_0$ are reference temperature and salinity ($T_0$ = 10$^{\circ}$C, $S_0$ = 35 g kg^-1^). Let's create a T-S diagram with this linearlized version of the equation of state.

```{r, fig.width = 5, fig.height = 5, eval = TRUE, echo = FALSE}
# Define ranges
T <- seq(0, 30, by = 0.1) # y-axis (temperature)
S <- seq(30, 40, by = 0.1) # x-axis (salinity)

fun.rho <- function(T, S) {
  rho_0 <- 1025 # reference sea water density in kg/m3
  alpha <- 1.5 * 10^(-4) # deg^{-1}
  beta <- 8 * 10^(-4) # psu ^{-1}
  T0 <- 10 # reference temperature (C)
  S0 <- 35 # reference salinity (g/kg)

  rho <- rho_0 * (1 - alpha * (T - T0) + beta * (S - S0))
  return(rho)
}


# Compute density using outer
# outer(T, S) -> rows = T, cols = S
rho_matrix <- outer(T, S, fun.rho)

# Transpose so contour interprets correctly
rho_matrix <- t(rho_matrix)

# Plot
contour(
  x = S, y = T, z = rho_matrix,
  xlab = "Salinity (g/kg)",
  ylab = "Temperature (°C)",
  main = "Linearlized T-S Diagram",
  levels = seq(500, 2000, by = 2),
  col = "blue", lwd = 1, labcex = 1.0
)

box.HL <- c(34, 2) # salinity and temperature of high-latitude box
box.LL <- c(36, 25) # salinity and temperature of high-latitude box
points(x = box.HL[1], y = box.HL[2], col = "blue", pch = 16) # high-latitude
points(x = box.LL[1], y = box.LL[2], col = "red", pch = 16) # low-latitude
grid()
```

You can see that the density values are similar, but not identical. However, using a linearlized equation of state keeps the math below more simple. Let's now finally get to the Stommel model.

### Phase Diagram

The Stommel model consists of two ODEs, which represent the non-dimensional salinity contrast ($x$), non-dimensional temperature contrast ($y$) (page 227 in respective paper):

$$ \frac{dx}{d\tau} = \delta(1-x) - \frac{x}{\lambda} |-y +Rx| $$
$$ \frac{dy}{d\tau} = 1 - y - \frac{y}{\lambda}|-y + Rx|$$

To summarize, the Stommel model has two variables ($x$ and $y$) and three main parameters ($R$, $\delta$, and $\lambda$). Let's define each variable and parameter:

* $x$ (-) is the non-dimensional salinity contrast between box 1 (polar) and box 2 (tropical): $x = \frac{\Delta S}{\Delta S^*}$, where $\Delta S^*$ is the typical salinity contrast between both boxes. In the model, $S^*$ denotes the salinity of the imposed boundary conditions. If $x<1$ then that means that the contrast is smaller than the imposed contrast.

* $y$ (-) is the non-dimensional temperature contrast between box 1 (polar) and box 2 (tropical): $y = \frac{\Delta T}{\Delta T^*}$, where $\Delta T^*$ is the typical temperature contrast between both boxes. In the model, $T^*$ denotes the salinity in the adjacent ocean basin. If $y<1$ then that means that the contrast is smaller than the imposed contrast.

* $\delta$ (-) is the ratio of the salinity transfer coefficient ($d$, in s^-1^) and temperature transfer coefficient ($c$, in s^-1^): $\delta = \frac{d}{c}$. If $\delta<1$ then the temperature transfer rate exceeds the salinity transfer rate.

* $\tau$ (-) is the non-dimensional time variable: $\tau = ct$ 

* $R$ (-) is a ratio that expresses the relative effect of salinity and temperature on seawater density: $R = \frac{\beta \Delta S^*}{\alpha \Delta T^*}$

* $\alpha$ (K^-1^) is the thermal contraction ($\alpha = 1.5 \times 10^{-4}$ K^-1^)

* $\beta$ (psu^-1^) is the saline expansion ($\beta = 8 \times10^{-4}$ psu^-1^)

* $\lambda$ (-) measures the relative strength of flow resistance to thermal forcing: $\lambda = \left( \frac{c}{4 \rho_0 \alpha \Delta T^*} \right) k$. The larger $\lambda$, the weaker the circulation response to temperature contrast.

* $k$ (kg m$^{-3}$ s) is the resistance of the capillary tube: $q=\frac{1}{k}(\rho_1-\rho_2)$

* $q$ (s$^{-1}$) is the flow rate: $q=\frac{1}{k}(\rho_1-\rho_2)$

Let us begin by plotting the free solution of the dynamical system for $R$ = 2, $\delta$ = 1/6, and $\lambda$ = 1/5, as well as initial conditions of $x = 0.1$ and $y = 0.1$:

```{r, echo = FALSE, fig.width = 10, fig.height = 5}
# Load required package
library(deSolve)

# Define parameters (change these to explore different behavior)
R <- 2
delta <- 1 / 6
lambda <- 1 / 5

# Define the system of ODEs
system <- function(t, state, parameters) {
  x <- state[1]
  y <- state[2]

  dx <- delta * (1 - x) - x / lambda * abs(-y + R * x)
  dy <- 1 - y - y / lambda * abs(-y + R * x)

  list(c(dx, dy))
}

# Initial conditions
initial_conditions <- c(0.9, 0.9)

# Time span
dt <- 0.01
times <- seq(0, 20, by = dt)

# Solve ODEs for each initial condition
solutions <- ode(
  y = initial_conditions,
  times = times,
  func = system,
  parms = NULL,
  method = "lsoda"
)

# Plot
df <- data.frame(solutions)
colnames(df) <- c("time", "x", "y")

par(mfrow = c(1, 2))

plot(
  x = df$time,
  y = df$x,
  type = "l",
  ylim = c(0, 1),
  ylab = "Non-dimensional Salinity Gradient",
  xlab = "Non-dimensional Time",
  main = "Salinity"
)
grid()

plot(
  x = df$time,
  y = df$y,
  type = "l",
  ylim = c(0, 1),
  ylab = "Non-dimensional Temperature Gradient",
  xlab = "Non-dimensional Time",
  main = "Temperature"
)
grid()
```

Rerun the code from the previous figure for initial conditions of $x = 0.1$ and $y = 0.1$.

```{r, fig.width = 10, fig.height = 5}
```

```{r, fig.width = 10, fig.height = 5, include = FALSE}
# Load required package
library(deSolve)

# Define parameters (change these to explore different behavior)
R <- 2
delta <- 1 / 6
lambda <- 1 / 5

# Define the system of ODEs
system <- function(t, state, parameters) {
  x <- state[1]
  y <- state[2]

  dx <- delta * (1 - x) - x / lambda * abs(-y + R * x)
  dy <- 1 - y - y / lambda * abs(-y + R * x)

  list(c(dx, dy))
}

# Initial conditions
initial_conditions <- c(0.1, 0.1)

# Time span

dt <- 0.01
times <- seq(0, 20, by = dt)

# Solve ODEs for each initial condition
solutions <- ode(
  y = initial_conditions,
  times = times,
  func = system,
  parms = NULL,
  method = "lsoda"
)

# Plot
df <- data.frame(solutions)
colnames(df) <- c("time", "x", "y")

par(mfrow = c(1, 2))

plot(
  x = df$time,
  y = df$x,
  type = "l",
  ylim = c(0, 1),
  ylab = "Non-dimensional Salinity Gradient",
  xlab = "Non-dimensional Time",
  main = "Salinity"
)
grid()

plot(
  x = df$time,
  y = df$y,
  type = "l",
  ylim = c(0, 1),
  ylab = "Non-dimensional Temperature Gradient",
  xlab = "Non-dimensional Time",
  main = "Temperature"
)
grid()
```

**Q: How does the change in initial conditions affect the equilibrium solution of the salinity contrast?**

* (A): The equilibrium salinity contrast increases to a value between 0.7 and 0.8
* (B): The equilibrium salinity contrast decreases to a value between 0.1 and 0.2
* (C): The equilibrium salinity contrast remains the same

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "B"
record_answerKey(answerKey)
```


**Q: Compare the four figures above with Figure 7 in Stommel's 1961 paper. Are your figures above consistent with what you see in Stommel's Figure 7.**

* (A): No, I only found 2 equilibrium solutions, while Stommel's figure finds 3 equilibrium solutions.
* (B): No, my equilibrium solutions don't match those from Stommel.
* (C): Yes, depending what initial conditions I choose I will end up at equilibrium point (a) or (c).
* (D): Yes, depending what initial conditions I choose I will end up at equilibrium point (b) or (c).

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "C"
record_answerKey(answerKey)
```

We can use the same function to plot non-dimensional salinity contrast ($x$) against non-dimensional temperature contrast ($y$) for different initial conditions. This leads to a so-called phase portrait, and corresponds to Stommel's Figure 7.

```{r, fig.width = 5, fig.height = 5, eval = TRUE, echo = FALSE}
# Define parameters (change these to explore different behavior)
R <- 2
delta <- 1 / 6
lambda <- 1 / 5

# Define the system of ODEs
system <- function(t, state, parameters) {
  x <- state[1]
  y <- state[2]

  dx <- delta * (1 - x) - x / lambda * abs(-y + R * x)
  dy <- 1 - y - y / lambda * abs(-y + R * x)

  list(c(dx, dy))
}

# Initial conditions
num <- seq(0, 1, 0.2)
zer <- rep(0, length(num))
one <- rep(1, length(num))

x <- c(zer, num, one, num)
y <- c(num, one, num, zer)


# Time span
dt <- 0.01
times <- seq(0, 20, by = dt)

n <- length(x)

result <- vector("list", n) # preallocate

for (i in 1:n)
{
  x.i <- x[i]
  y.i <- y[i]

  initial_conditions <- c(x.i, y.i)

  # Solve ODEs for each initial condition

  solutions <- ode(
    y = initial_conditions,
    times = times,
    func = system,
    parms = NULL,
    method = "lsoda"
  )

  df <- data.frame(solutions[, 2], solutions[, 3])
  colnames(df) <- c("x", "y")
  result[[i]] <- df
}

x <- result[[1]]$x
y <- result[[1]]$y


plot(x, y,
  type = "l", xlim = c(0, 1), ylim = c(0, 1),
  xlab = "Non-dimensional salt contrast",
  ylab = "Non-dimensional temperature contrast"
)

for (i in 1:n) {
  x <- result[[i]]$x
  y <- result[[i]]$y

  lines(x, y, type = "l")
}
grid()
```

**Q: Which of the points below corresponds to an equilibrium point?**

* (A): x = 0.4, y = 0.6
* (B): x = 0.4, y = 0.8
* (C): x = 0.2, y = 1.0
* (D): x = 0.2, y = 0.0

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "B"
record_answerKey(answerKey)
```

The phase portrait above identified three equilibrium points. Let us know calculate those equilibrium points from the equations. In the section on linear algebra, we did this by setting the ODEs to zero and solving for each variable. For more complicated ODEs, we can also approximate the equilibrium points numerically, which is what we will do here.

```{r, eval = TRUE, echo = FALSE}
library(rootSolve)

R <- 2
delta <- 1 / 6
lambda <- 1 / 5

# system of equations
f <- function(z) {
  x <- z[1]
  y <- z[2]

  dx <- delta * (1 - x) - x / lambda * abs(-y + R * x)
  dy <- 1 - y - y / lambda * abs(-y + R * x)

  c(dx, dy)
}

# grid of starting guesses
xstart <- seq(-1, 2, length = 8)
ystart <- seq(-1, 2, length = 8)

solutions <- list()

for (xs in xstart) {
  for (ys in ystart) {
    out <- try(multiroot(f, start = c(xs, ys)), silent = TRUE)
    if (!inherits(out, "try-error")) {
      solutions[[length(solutions) + 1]] <- out$root
    }
  }
}

# combine and remove duplicates
solmat <- do.call(rbind, solutions)

unique_solutions <- unique(round(solmat, 4))
print(unique_solutions)
```

Let's plot these three equilibrium points into our phase portrait. Do this by copying the code from above and then add ```points(x = 0.1350, y = 0.4836), pch = 16, col = "red"``` below. Do this for all three points.

```{r, fig.width = 5, fig.height = 5}
```


```{r, fig.width = 5, fig.height = 5, echo = FALSE}
# Define parameters (change these to explore different behavior)
R <- 2
delta <- 1 / 6
lambda <- 1 / 5

# Define the system of ODEs
system <- function(t, state, parameters) {
  x <- state[1]
  y <- state[2]

  dx <- delta * (1 - x) - x / lambda * abs(-y + R * x)
  dy <- 1 - y - y / lambda * abs(-y + R * x)

  list(c(dx, dy))
}

# Initial conditions
num <- seq(0, 1, 0.2)
zer <- rep(0, length(num))
one <- rep(1, length(num))

x <- c(zer, num, one, num)
y <- c(num, one, num, zer)


# Time span
dt <- 0.01
times <- seq(0, 20, by = dt)

n <- length(x)

result <- vector("list", n) # preallocate

for (i in 1:n)
{
  x.i <- x[i]
  y.i <- y[i]

  initial_conditions <- c(x.i, y.i)

  # Solve ODEs for each initial condition

  solutions <- ode(
    y = initial_conditions,
    times = times,
    func = system,
    parms = NULL,
    method = "lsoda"
  )

  df <- data.frame(solutions[, 2], solutions[, 3])
  colnames(df) <- c("x", "y")
  result[[i]] <- df
}

x <- result[[1]]$x
y <- result[[1]]$y


plot(x, y,
  type = "l", xlim = c(0, 1), ylim = c(0, 1),
  xlab = "Non-dimensional salt contrast",
  ylab = "Non-dimensional temperature contrast"
)

for (i in 1:n) {
  x <- result[[i]]$x
  y <- result[[i]]$y

  lines(x, y, type = "l")
}
grid()

points(x = unique_solutions[1, 1], y = unique_solutions[1, 2], pch = 16, col = "red")
points(x = unique_solutions[2, 1], y = unique_solutions[2, 2], pch = 16, col = "blue")
points(x = unique_solutions[3, 1], y = unique_solutions[3, 2], pch = 16, col = "green")
```

### Jacobian matrix and equilibrium eigenvalues

Let's recall our set of ODEs:

$$ \frac{dx}{d\tau} = \delta(1-x) - \frac{x}{\lambda} |-y +Rx| $$
$$ \frac{dy}{d\tau} = 1 - y - \frac{y}{\lambda}|-y + Rx|$$

Recall the Jacobian matrix:

$$
\mathbf{J} =
\begin{bmatrix}
\frac{\partial f_1}{\partial x} & \frac{\partial f_1}{\partial y} \\
\frac{\partial f_2}{\partial x} & \frac{\partial f_2}{\partial y} \\
\end{bmatrix}
$$

Rather than calculating the derivatives analytically, we will approach them numerically:

```{r, echo = FALSE, message = FALSE}
library(numDeriv)

# constants

R <- 2
delta <- 1 / 6
lambda <- 1 / 5


my.ODEs <- function(z) {
  x <- z[1]
  y <- z[2]

  dx <- delta * (1 - x) - x / lambda * abs(-y + R * x)
  dy <- 1 - y - y / lambda * abs(-y + R * x)

  c(dx, dy)
}

# Equilibrium solutions
zstar <- unique_solutions[1, ]
J <- jacobian(func = my.ODEs, x = zstar)
eigenvalues <- eigen(J)$values
eigen.01 <- eigenvalues

zstar <- unique_solutions[2, ]
J <- jacobian(func = my.ODEs, x = zstar)
eigenvalues <- eigen(J)$values
eigen.02 <- eigenvalues

zstar <- unique_solutions[3, ]
J <- jacobian(func = my.ODEs, x = zstar)
eigenvalues <- eigen(J)$values
eigen.03 <- eigenvalues

df <- data.frame(eigen.01, eigen.02, eigen.03)
print(df)
```

Recall that the equilibrium eigenvalues characterize the equilibrium point, e.g.:

* If all eigenvalues $>0$: unstable node
* If all eigenvalues $<0$: stable node
* If the sign varies among eigenvalues: saddle (unstable)
* If any $\lambda = 0$: tipping point

The second equilibrium point is described by a complex number with a real and an imaginary component. Here we are only concerned with the real component of the number. 

**Q: Classify each of the three equilibrium points**

* (A): unstable node, unstable saddle, stable node
* (B): stable node, unstable saddle, stable node
* (C): stable node, stable node, unstable saddle
* (D): stable node, stable node, unstable saddle

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "C"
record_answerKey(answerKey)
```

### Bifurcation diagram

Recall that $\lambda$ (-) measures the relative strength of flow resistance to thermal forcing:

$$\lambda = \left( \frac{c}{4 \rho_0 \alpha \Delta T^*} \right) k$$. The larger $\lambda$, the weaker the circulation response to temperature contrast. Let us now vary the parameter $\lambda$ and calculate the equilibrium points for different values of $\lambda$. This will result in the bifurcation diagram below.

```{r, fig.width = 10, fig.height = 5, eval = TRUE, echo = FALSE, warning  = FALSE}
library(rootSolve)

R <- 2
delta <- 1 / 6

f <- function(z, lambda) {
  x <- z[1]
  y <- z[2]

  dx <- delta * (1 - x) - x / lambda * abs(-y + R * x)
  dy <- 1 - y - y / lambda * abs(-y + R * x)

  c(dx, dy)
}

find_eq <- function(lambda) {
  xstart <- seq(-0.5, 1.5, length = 6)
  ystart <- seq(-0.5, 1.5, length = 6)
  sols <- list()

  for (xs in xstart) {
    for (ys in ystart) {
      out <- try(
        multiroot(f,
          start = c(xs, ys),
          lambda = lambda
        ),
        silent = TRUE
      )

      if (!inherits(out, "try-error")) {
        root <- out$root

        # --- convergence check ---
        res <- f(root, lambda)
        if (max(abs(res)) < 1e-6) { # tolerance
          sols[[length(sols) + 1]] <- root
        }
      }
    }
  }

  if (length(sols) == 0) {
    return(NULL)
  }
  unique(round(do.call(rbind, sols), 4))
}


# sweep lambda
lambdas <- seq(0.01, 0.4, by = 0.01)
results <- data.frame()

for (lam in lambdas) {
  eq <- find_eq(lam)
  if (is.null(eq)) next

  results <- rbind(
    results,
    data.frame(
      lambda = lam,
      x = eq[, 1],
      y = eq[, 2]
    )
  )
}

par(mfrow = c(1, 2))
# plot bifurcation diagram (x equilibrium)
plot(results$lambda, results$x,
  pch = 16,
  xlab = expression(lambda),
  ylab = "Equilibrium x",
  main = "(a) Salinity contrast (x)"
)
grid()
abline(h = 0.4, col = "blue")
abline(h = 0.26, col = "blue")

plot(results$lambda, results$y,
  pch = 16,
  xlab = expression(lambda),
  ylab = "Equilibrium y",
  main = "(b) Temperature contrast (y)"
)
grid()
abline(h = 0.8, col = "blue")
abline(h = 0.68, col = "blue")
```

**Q: The blue horizontal lines separate the stable form the unstable branches. Which are the stable and which are the unstable branches?**

* (A): top = unstable, middle = stable, bottom = unstable
* (B): top = stable, middle = stable, bottom = unstable
* (C): top = unstable, middle = stable, bottom = stable
* (D): top = stable, middle = unstable, bottom = stable

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "D"
record_answerKey(answerKey)
```

From the equilibrium $x$ (non-dimensional salinity contrast) and $y$ (non-dimensional temperature contrast) values, we can calculated the corresponding dimensionless flow rate $f$:

$$f = (-y + Rx) /\lambda $$

* If $f<0$ then the circulation is in a temperature-mode (T-mode): the deep branch flow is in the direction set by temperature contrast (from cold to warm = from high to low latitudes)

* If $f>0$ then the circulation is in a salinity mode (S-mode): the deep branch flow is in the direction set by salinity contrast (from salty to fresh = from low to high latitudes)

```{r, fig.width = 5, fig.height = 5, eval = TRUE, echo = FALSE, warning = FALSE}
library(rootSolve)

R <- 2
delta <- 1 / 6

f <- function(z, lambda) {
  x <- z[1]
  y <- z[2]

  dx <- delta * (1 - x) - x / lambda * abs(-y + R * x)
  dy <- 1 - y - y / lambda * abs(-y + R * x)

  c(dx, dy)
}

find_eq <- function(lambda) {
  xstart <- seq(-0.5, 1.5, length = 6)
  ystart <- seq(-0.5, 1.5, length = 6)
  sols <- list()

  for (xs in xstart) {
    for (ys in ystart) {
      out <- try(
        multiroot(f,
          start = c(xs, ys),
          lambda = lambda
        ),
        silent = TRUE
      )

      if (!inherits(out, "try-error")) {
        root <- out$root

        # --- convergence check ---
        res <- f(root, lambda)
        if (max(abs(res)) < 1e-6) { # tolerance
          sols[[length(sols) + 1]] <- root
        }
      }
    }
  }

  if (length(sols) == 0) {
    return(NULL)
  }
  unique(round(do.call(rbind, sols), 4))
}


# sweep lambda
lambdas <- seq(0.01, 0.4, by = 0.01)
results <- data.frame()

for (lam in lambdas) {
  eq <- find_eq(lam)
  if (is.null(eq)) next

  results <- rbind(
    results,
    data.frame(
      lambda = lam,
      x = eq[, 1],
      y = eq[, 2]
    )
  )
}

lambda <- results$lambda
x <- results$x
y <- results$y


f <- (-y + R * x) / lambda

par(mfrow = c(1, 1))
# plot bifurcation diagram (x equilibrium)
plot(results$lambda, f,
  pch = 16,
  xlab = expression(lambda),
  ylab = "Equilibrium f",
  main = "Dimensionless flow rate f"
)
grid()
abline(h = 0, col = "red")
```

**Q: Consider you are in the T-mode and $\lambda$ = 0.3. Let's say that $\lambda$ increases to 0.35. What will happen with the direction of the flow?**

* (A): The flow stops.
* (B): The direction of the flow remains the same.
* (C): The deep flow reverses direction, switching from equatorward to poleward transport.
* (D): The deep flow reverses direction, switching from poleward to equatorward transport.

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "C"
record_answerKey(answerKey)
```

**Q: Next, imagine $\lambda$ then returns from 0.35 back to 0.3. How will this affect the flow?**

* (A): The flow stops.
* (B): The direction of the flow remains the same.
* (C): The deep flow reverses direction, switching from equatorward to poleward transport.
* (D): The deep flow reverses direction, switching from poleward to equatorward transport.

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "B"
record_answerKey(answerKey)
```

**Q: Have another look at your phase diagram. Calculate $f$ for each point and evaluate which of the equilibrium points corresponds to the T-mode, and which to the S-mode?**

* (A): red (x=0.13, y = 0.48) = T-mode, blue (x = 0.43, y = 0.82) = S-mode, green (x = 0.35, y = 0.76) (T-mode)
* (B): red (x=0.13, y = 0.48) = S-mode, blue (x = 0.43, y = 0.82) = T-mode, green (x = 0.35, y = 0.76) (S-mode)
* (C): red (x=0.13, y = 0.48) = T-mode, blue (x = 0.43, y = 0.82) = S-mode, green (x = 0.35, y = 0.76) (S-mode)
* (D): red (x=0.13, y = 0.48) = S-mode, blue (x = 0.43, y = 0.82) = S-mode, green (x = 0.35, y = 0.76) (T-mode)

```{r, include = FALSE}
R <- 2
delta <- 1 / 6
lambda <- 1 / 5

x <- unique_solutions[1, 1]
y <- unique_solutions[1, 2]
f1 <- (-y + R * x) / lambda # red
print(f1)

x <- unique_solutions[2, 1]
y <- unique_solutions[2, 2]
f2 <- (-y + R * x) / lambda # blue
print(f2)

x <- unique_solutions[3, 1]
y <- unique_solutions[3, 2]
f3 <- (-y + R * x) / lambda # green
print(f3)
```

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "A"
record_answerKey(answerKey)
```

## CLIMBER-X: Hosing experiments

We are now going to look at the hosing experiment conducted with CLIMBER-X. To conduct the hosing experiment, we run to he following line of code

```{bash, eval=FALSE, echo=TRUE}
./runme -rs -q medium -w 60:00:00 --omp 32 -o $output/$outdir/hyst2050/up 
-p ctl.nyears=30000 ocn.l_hosing=T ocn.hosing_ini=-0.3 ocn.hosing_trend=0.02
ocn.lat_min_hosing=20 ocn.lat_max_hosing=50
```

Let's assess what this means:

* ```./runme -rs -q medium -w 60:00:00``` runs the model and tells the HPC how much computational resources the model requires

* ```-o $output/$outdir/hyst2050/up``` tells the HPC where to store the output

* ```-p ``` sets the parameters. The meaning of each parameter is outlined in the different namelist files (```.nml```)located in ```/CLIMBER-X/climber-x/nml/```

*  You can find the meaning of each parameter as follows:

```{bash, eval=FALSE, echo=TRUE}
# go to the directory where the namelist files are stored
cd /CLIMBER-X/climber-x/nml/

# search for the parameter in all namelist files:
grep -r "nyears" *.nml
```

Here a list of all the parameters that were set for this experiment:

* ```ctl.nyears=30000```: number of simulation years

* ```ocn.l_hosing=T```: apply additional freshwater hosing? ($T$ stands for TRUE, $F$ for FALSE)

* ```ocn.hosing_ini=-0.3```: Sv, inital/constant value of hosing flux

* ```ocn.hosing_trend=0.02```: Sv/kyr, rate at which hosing flux is increased/decreased 

* ```ocn.lat_min_hosing=20```: lower latitude bound for region of freshwater flux

* ```ocn.lat_max_hosing=50```: higher latitude bound for region of freshwater flux

The command to run the freshwater hosing experiment only mentions a few parameters. For all other parameters, the default setting is chosen. For instance, the atmospheric CO~2~ concentration is set to a constant value of 280 ppmv. Recall that Sv = Sverdrup = 10^6^ km^2^ s^-1^. Let's now translate the command below into words. Again:

```{bash, eval=FALSE, echo=TRUE}
./runme -rs -q medium -w 60:00:00 --omp 32 -o $output/$outdir/hyst2050/up 
-p ctl.nyears=30000 ocn.l_hosing=T ocn.hosing_ini=-0.3 ocn.hosing_trend=0.02
ocn.lat_min_hosing=20 ocn.lat_max_hosing=50
```

This means: simulate 30,000 years starting with an initial freshwater input of -0.3 Sv  between 20 and 50 degrees latitude in the North Atlantic. The freshwater flux is increased by a rate of 0.02 Sv/1000 years, causing it to increase from -0.3 Sv to +0.3 Sv after 30,000 years. What does a negative freshwater flux mean? A negative freshwater flux means that freshwater is taken out of the ocean (evaporation > precipitation), increasing the seawater's salinity. 

**Q: Reproduce the figure below to illustrate how the hosing changes over time**

```{r, fig.width = 5, fig.height = 5}
```

```{r, fig.width = 5, fig.height = 5, eval = TRUE, echo = FALSE}
time <- seq(1, 30000, 1)
rate <- 0.02 / 1000 # change in Sv per 1000 years
intercept <- -0.3 # Sv
hosing <- rate * time + intercept

plot(x = time, y = hosing, main = "Freshwater Hosing between 20-50 Deg Latitude", xlab = "Time (years)", ylab = "Freshwater hosing (Sv)", type = "l")
grid()
```

We have also submitted another simulation:

```{bash, eval=FALSE, echo=TRUE}
./runme -rs -q medium -w 60:00:00 --omp 32 -o $output/$outdir/hyst2050/down 
-p ctl.nyears=30000 ocn.l_hosing=T ocn.hosing_ini=0.3  ocn.hosing_trend=-0.02 
ocn.lat_min_hosing=20 ocn.lat_max_hosing=50
```

**Q: Make a figure similar to the one above that illustrates how the freshwater hosing changes over time.**

```{r}
```


### Hysteresis

The AMOC strength is defined as the maximum of the Atlantic meridional streamfunction deeper than 700 m. The initial condition for both these experiments (up, and down) is a pre-industrial equilibrium simulation run for 10 000 years with 280 ppm of CO2 and present-day ice sheets. Let's now visualize the results from both experiments.


```{r, fig.width = 5, fig.height = 5, warning = FALSE, echo = FALSE, message = FALSE}
library(clover)
library(ncdf4)
library(fields)
library(viridis)

dataDir <- "/global/home/hpc5112/output/bench_v1.4.0/hyst2050/down"
ncFileName <- "ocn_ts.nc"
variableName <- "amoc26N"

# down experiment
dataDir <- "/global/home/hpc5112/output/bench_v1.4.0/hyst2050/up"
ncFileName <- "ocn_ts.nc"
variableName <- "amoc26N"

data <- clover::clover.data(dataDir, ncFileName, variableName)

# summary(data)

up <- data$values
time <- data$time
long_name <- data$long_name
units <- data$units

# Hosing up
time <- time - time[1] + 1
rate <- 0.02 / 1000 # change in Sv per 1000 years
intercept <- -0.3 # Sv
hosing <- rate * time + intercept
hosing.up <- hosing

xlab <- "Freshwater hosing (Sv)"
ylab <- paste0(variableName, " (", units, ")")

# down experiment
dataDir <- "/global/home/hpc5112/output/bench_v1.4.0/hyst2050/down"
data <- clover::clover.data(dataDir, ncFileName, variableName)
down <- data$values
time <- data$time

# Hosing down
time <- time - time[1] + 1
rate <- -0.02 / 1000 # change in Sv per 1000 years
intercept <- 0.3 # Sv
hosing <- rate * time + intercept
hosing.down <- hosing

par(mar = c(5, 5, 3, 3))
plot(
  x = hosing.up, y = up,
  main = long_name,
  type = "l", col = "red",
  xlab = xlab, ylab = ylab, ylim = c(0, 35), xlim = c(-0.1, 0.25)
)
lines(x = hosing.down, y = down, col = "blue")
legend("topright", c("up", "down"), pch = 16, col = c("red", "blue"), bty = "n")
grid()
```

**Q: Interpret the figure above**

* (A): A reduction in salinity weakens the AMOC and may cause it to collapse beyond a tipping point. However, once salinity is increased again, the AMOC immediately returns to its original strength at exactly the same threshold at which it collapsed.

* (B): A reduction in salinity weakens the AMOC and can ultimately trigger its collapse once a critical tipping point is crossed. If salinity is subsequently increased, the AMOC does not immediately recover its original strength. Instead, salinity must be restored well beyond the initial threshold before the AMOC returns to its previous state.

* (C): A decrease in salinity strengthens the AMOC up to a critical threshold, beyond which the circulation abruptly collapses. Increasing salinity afterward further delays recovery of the system.

* (D): When salinity declines past a tipping point, the AMOC collapses. Once salinity is increased again, the AMOC gradually strengthens but stabilizes at a permanently weaker state, even if salinity fully returns to its initial level.

```{r, echo = FALSE}
answer <- "Add your answer here" # (e.g. "A")
record_answer(answer)
print(answer)
```

```{r, include = FALSE}
answerKey <- "B"
record_answerKey(answerKey)
```

Let's plot how the AMOC looks like when it is on (high salinity) and off (low salinity)

```{r, fig.width = 10, fig.height = 5, eval = TRUE, echo = FALSE}
# down experiment
dataDir <- "/global/home/hpc5112/output/bench_v1.4.0/hyst2050/up"
ncFileName <- "ocn.nc"
variableName <- "opsi_a"

# Read in data
data <- clover::clover.data(dataDir, ncFileName, variableName)

# Have a look at the content
# summary(data)

# Look at the dimensions
# print(data$dimension_names)

# Extract values
dim_values <- data$dimension_values
values <- data$values
time <- data$time

# Choose years for plotting
data.yr <- values[, , 13, 6:8] # year 6000 - 8000
data.mean <- apply(data.yr, MARGIN = c(1, 2), FUN = mean, na.rm = TRUE)
data.mean <- data.mean[, ncol(data.mean):1]
data.mean[data.mean == 0] <- NA
data.salt <- data.mean

data.yr <- values[, , 13, 25] # year 21000-24000
data.mean <- apply(data.yr, MARGIN = c(1, 2), FUN = mean, na.rm = TRUE)
data.mean <- data.mean[, ncol(data.mean):1]
data.mean[data.mean == 0] <- NA
data.fresh <- data.mean

latv1 <- dim_values$latv1
levw <- dim_values$levw
levw <- (-1) * rev(levw)

zlim <- range(data.salt, na.rm = TRUE)


my.title.salt <- paste0(data$long_name, " (", data$units, ") \n", "(high salinity)")
my.title.fresh <- paste0(data$long_name, " (", data$units, ") \n", "(low salinity)")

par(mfrow = c(1, 2), mar = c(4, 4, 4, 4))

fields::image.plot(latv1, levw, data.salt,
  xlim = c(-30, 80),
  zlim = zlim,
  col = viridis::viridis(50),
  xlab = "Latitude", ylab = "Depth (m)",
  main = my.title.salt
)
grid()

contour(latv1, levw, data.salt, add = TRUE)

fields::image.plot(latv1, levw, data.fresh,
  xlim = c(-30, 80),
  zlim = zlim,
  col = viridis::viridis(50),
  xlab = "Latitude", ylab = "Depth (m)",
  main = my.title.fresh
)
grid()
contour(latv1, levw, data.fresh, add = TRUE)
```

Let us now plot the AMOC strength at 26 degrees northern latitude as a function of depth, where each line represents an increase in freshwater hosing.

```{r, fig.width = 5, fig.height = 5, eval = TRUE, echo = FALSE}
# down experiment

dataDir <- "/global/home/hpc5112/output/bench_v1.4.0/hyst2050/up"
ncFileName <- "ocn.nc"
variableName <- "opsi_a"

data <- clover::clover.data(dataDir, ncFileName, variableName)
# summary(data)

dim_values <- data$dimension_values

levw <- dim_values$levw
levw <- (-1) * rev(levw)

dim_values <- data$dimension_values

values <- data$values
# data$dimension_names
# data$dimension_sizes
time <- data$time

latitude <- 26

data.lat <- values[26, , 13, ]

my.col <- rev(viridis::viridis(ncol(data.lat)))

my.xlab <- paste0(data$long_name, " (", data$units, ") \n", "(26 N Latitude)")

par(mfrow = c(1, 1), mar = c(4, 4, 4, 4))

matplot(
  x = data.lat, y = levw,
  type = "l",
  lty = 1,
  col = my.col,
  xlab = my.xlab,
  ylab = "Depth (m)",
  main = "AMOC at 26N Latitude"
)
grid()
```

### Reflection

**Q: Write a concise paragraph summarizing the main mechanisms or concepts you learned in this tutorial. Focus on explaining how they are connected.**

* Answer: ___

**Q: Which concept from this tutorial do you feel most confident explaining to someone else? Briefly state why.**

* Answer: ___

**Q: Which concept remains least clear to you? Be specific about what part is confusing.**

* Answer: ___

```{r, include=FALSE}
tutorialName <- "ocean-circulation"

output <- list(
  student_id = student_id,
  tutorial   = tutorialName,
  timestamp  = Sys.time(),
  answers    = answers
)

filename <- paste0("answers_", tutorialName, "_", student_id, ".json")
jsonlite::write_json(output, filename, pretty = TRUE)
rm(output)
```

```{r generate_solution_JSON, include=FALSE}
# Generate solution JSON file for this tutorial

library(jsonlite)

# Build the answer key object
output <- list(
  tutorial = tutorialName,
  generated_at = Sys.time(),
  answers = answerKeys
)

# Write JSON file
filename <- paste0("answerKey_", tutorialName, ".json")
jsonlite::write_json(output, filename, pretty = TRUE)
```


<!--chapter:end:02-ocean-circulation.Rmd-->

